---
title: "MB5370:Assesment: End-to-end analysis in R"
author: "Jacqueline Hintz"
date: "2024-05-16"
output:
  pdf_document: default
  html_document: default
---

First Went into QFISH website and downloaded Shark Control Program, and selected the number caught by area, calendar year and species group. 
This data set was a complete mess when downloading it into a csv and opening it in excel. The table has the years along the top if the table and we need them on the sides. We need to pivot the columns and rows to make it tidy

In excel I tidied some of the columns, and rows. I deleted the grand total and fixed some random little things to make sure R can read it.

#Imported QFISH dataset
```{r}
library(tidyverse)
library(readr)
QFISH <- read_csv("data/QFISH.csv")
QFISH <- mutate_all(QFISH, ~replace(., is.na(.),0)) # replaced all the N/As with zeros
```

# Tidying Data

```{r}
QFISH <- read_csv("data/QFISH.csv") # have R read the tidied csv from excel that was imported
```


#Now we want to make the rows the columns like catch and calender year 
#Starting with year 2001. Repeated these steps from 2002-2005.
```{r}
yr2001 <- QFISH |> # selected year 2001 first to start with
  dplyr::select(Area, starts_with("2001")) # if errors here put yr2001 infront of Area then comma,

colnames(yr2001) <- unlist(yr2001[1,]) # removes repeated row  (first row- year)

yr2001 <-
  yr2001 |>
  slice(-1) |> #removes repeated catch row
    rename(Area = Catch) |> #replaces the heading catch with area
  mutate(Year = 2001) #adds the row year and fills with 2001
```

Next year 2002
```{r}
yr2002 <- QFISH |> 
  select(Area, starts_with("2002"))

colnames(yr2002) <- unlist(yr2002[1,]) 

yr2002 <-
  yr2002 |>
  slice(-1) |> 
  rename(Area = Catch) |> 
  mutate(Year = 2002)
```

Then 2003
```{r}
yr2003 <- QFISH |>
  select(Area, starts_with("2003"))

colnames(yr2003) <- unlist(yr2003[1,])
 
yr2003 <- 
  yr2003 |> 
  slice(-1) |>
  rename(Area = Catch) |>
  mutate(Year = 2003)
```



Then 2004
```{r}
yr2004 <- QFISH |> 
  select(Area, starts_with("2004")) 

colnames(yr2004) <- unlist(yr2004[1,]) 

yr2004 <-
  yr2004 |>
  slice(-1) |> 
  rename(Area = Catch) |> 
  mutate(Year = 2004)
```

And finally 2005. I know this is not the best solution for this large data set that went all the way to 2023, but it the best one I could come up with. 
```{r}
yr2005 <- QFISH |>
  select(Area, starts_with("2003"))

colnames(yr2005) <- unlist(yr2005[1,])
 
yr2005 <- 
  yr2005 |> 
  slice(-1) |>
  rename(Area = Catch) |>
  mutate(Year = 2005)
```


Now we need to join the tables. row binding. 
```{r}
QFISH_new <- 
  bind_rows(yr2001,yr2002, yr2003, yr2004, yr2005) |> 
  mutate(Area = as.factor(Area)) |> #Convert the Area variable to factor ( will be chr otherwise and will not graphg)this just a way to make Area not be affected by next line of code
  mutate_if(is.character, as.numeric)  # The other vars (apart from 'Year') are  chr so  won't be recognised as numeric if plotted on continuous X or Y axis, so we'll convert
```


#checking the structure od data before plotting
```{r}
glimpse(QFISH_new) # Exactly as required now)
```

Now graph the new QFISH Data.
```{r}
ggplot(data = QFISH_new) +
  geom_point(mapping = aes(x = Year, y = Total, colour = Area)) + 
  theme_classic()
```
```{r}
ggplot(data = QFISH_new) +
  geom_point(mapping = aes(x = Year, y = Total, colour = Area)) +
  scale_y_continuous(breaks = c(0,200), limits = c(0,250)) +
  scale_x_continuous() +
  theme_classic()
```

#Final gragh with out the N/A on y-axis




