---
title: "MB5370:Assesment: End-to-end analysis in R"
author: "Jacqueline Hintz"
date: "2024-05-16"
output: html_document
---

Went into Qish website.

picked the Shark Program nu,ber by caught by area, calender year and species group dataset that offers a data tbale by region, year and species. 

#Imported QFISH dataset

library(readr)
> QFISH <- read_csv("data/QFISH.csv")

# Tidying Data
The table has the years along the top if the table and we need them on the sides. We need to pivot the columns and rows to make it tidy


In excel I tidy some of the columns, and rows. I deleted the grand total and fixed some random little things to make sure R can read it.

QFISH <- read_csv("data/QFISH1.csv")


Now we want to make the rows the columns like catch and calender year 

```{r}
yr2001 <- QFISH |> # selected year 2001 first to start with
  select(Area, starts_with("2001")) # if errors here put yr2001 infront of Area then comma

colnames(yr2001) <- unlist(yr2001[1,]) # removes repeated row  (first row- year)

yr2001 <-
  yr2001 |>
  slice(-1) |> # removes catch row?
  rename(Area = Catch) |> # replaced Catch header with Area
  mutate(Year = 2001) # added a year colum
```

```{r}
yr2002 <- QFISH |> # selected year 2002 first to start with
  select(Area, starts_with("2002"))

colnames(yr2002) <- unlist(yr2002[1,]) # removes repeated row  (first row -year)

yr2002 <-
  yr2002 |>
  slice(-1) |> # removes catch row?
  rename(Area = Catch) |> # replaced Catch header with Area
  mutate(Year = 2002)
```

```{r}
yr2003 <- QFISH |>
  select(Area, starts_with("2003"))

colnames(yr2003) <- unlist(yr2003[1,])
 
yr2003 <- 
  yr2003 |> 
  slice(-1) |>
  rename(Area = Catch) |>
  mutate(Year = 2003)
```


Now we need to join the tables. row bind

```{r}
QFISH_new <- 
  bind_rows(yr2001,yr2002, yr2003)
```








