---
title: "Workshop4_Spatial_data_in_r"
author: "Jacqueline Hintz"
date: "2024-05-17"
output: html_document
---

### Workshop 4: Spatial data inR

#Install packages

```{r}
# install and load your packages
install.packages("sf") 
install.packages("terra")
install.packages("tmap")


#load into R library
library(tidyverse)
library(sf) # simple features
library (terra) # for raster
library(tmap) # Thematic maps are geographical maps in which spatial data distributions are visualized

```

#task now is - using R - to make a map of copepods in relation to temperature
```{r}
#load the copepod data into R studio
library(readr)
dat <- read_csv("data-for-course/copepods_raw.csv")
dat
```

# notice here the  silk_id column, which is just the ID for each of the silks, onto which plankton are recorded

# Data exploration
```{r}
library(ggplot2)
ggplot(dat) + 
  aes(x = longitude, y = latitude, color = richness_raw) +
  geom_point()
```

#it should show the location of every segment and color the points by species richness. notice that the x and y. It is not a map!

#now lets look at data richness
```{r}
ggplot(dat, aes(x = latitude, y = richness_raw)) + 
  stat_smooth() + 
  geom_point()
```

#Something look odd with this graph. Unnatural pattern at about latitude -40. This would be the point in time to talk with your collaborator to try and work out whts going on,

#Getting going with maps
#First, we will turn our point data into a spatially referenced data frame using the sf package (sf stands for ‘simple features’) which is an open standard for geospatial databases. For those that think in GIS, you can think of this format as a shapefile or feature collection.

```{r}
sdat <- st_as_sf(dat, coords = c("longitude", "latitude"), crs = 4326)
```

# this turned our data into simple feature collection
#As is good practice (that I hope you’re learned by now!) use ?st_as_sf to see what else it can convert and what all these arguments mean.
#st_as_sf converts different data types to simple features. 
#dat is our original data. 
#coords gives the names of the columns that relate to the spatial coordinates (in order of X coordinate followed by Y coordinate).
#crs stands for coordinate reference system which we will discuss next.


#Coordinate referance system: is a short- hand code for a specific coordinate referance system (CRS)
#crs we are using is crs.  For example, for the EPSG code 4326 we can find out: 1) what the name of this crs is, 2) the corresponding proj4string, and 3) the WKT (WKT is a string that contains all of the detailed info we need to define a crs)

```{r}
crs4326 <- st_crs(4326)
crs4326 #look at the whole CRS
crs4326$Name #pull out just the name of the crs
"WGS 84"
crs4326$wkt # crs in well-known text format
```


```{r}
sdat
```

#Cartography
```{r}
plot(sdat["richness_raw"])
```



```{r}
plot(sdat) #creates panel for every variable on the dataframe.
```


#Thematic maps for communication

```{r}
#using tmap

map <- tm_shape(sdat) +
  tm_dots(col = "richness_raw")
```

#note the tm_dots to plot dots of the cordinates. othe options are tm_polygons, we have chosen "richness_raw" as the color scale 
#use tmap_save to save the map to your working dirc. 

```{r}
tmap_save(map, filename = "Richeness-map.png", width = 600, height = 600)
```

#Mapping spatial polygons as layers
#Sf package can handle many types of spatail data, inclduing shapes like polygons. To practice with polygons we will load in a map of AUS and a map of AUS's continential shelf using tmap to add to these layers.

#loading shapfiles
#Unlike the data we just mapped, which was a .csv file with coordinate columns, the polygons in this copepod data are stored as shapefiles. Note that .shp files are generally considered an undesirable file format because they are inefficient at storing data and to save one shapefile you actually create multiple files. This means bits of the file might be lost if you transfer the data somewhere else. Even in GIS software these days, we are moving well away from shapefiles to use other data formats.A better format than shapefile is the Geopackage which can save and compress multiple different data types all in a single file.
#We are working with shapefiles in this case study because it is still the most likely format you’ll encounter when someone sends you a spatial dataset, but I encourage you to save your personal data in the .gpkg format as you move forward. We can read shapefiles directly into R with the st_read command (which is like read_csv, but for spatial files):

```{r}
aus <- st_read("data/data-for-course/spatial-data/Aussie/Aussie.shp")
```

#output:Reading layer `Aussie' from data source 
#`C:\Users\jacqu\OneDrive\Documents\BC5101_ #Advanced_Genetics_2024\Module_4_Data_science_R\Mod_4_personal\data\data-for-course\spatial-data\Aussie\Aussie.shp' 
#####using driver `ESRI Shapefile'
#Simple feature collection with 8 features and 1 field
#Geometry type: MULTIPOLYGON
#Dimension:     XY
#Bounding box:  xmin: 112.9211 ymin: -43.63192 xmax: 153.6389 ymax: -9.229614
#Geodetic CRS:  WGS 84

```{r}
shelf <- st_read("data-for-course/spatial-data/aus_shelf/aus_shelf.shp")
```
#output: Reading layer `aus_shelf' from data source 
##`C:\Users\jacqu\OneDrive\Documents\BC5101_ #Advanced_Genetics_2024\Module_4_Data_science_R\Mod_4_personal\data-for-course\spatial-data\aus_shelf\aus_shelf.shp' 
#  using driver `ESRI Shapefile'
#Simple feature collection with 1 feature and 1 field
#Geometry type: MULTIPOLYGON
#Dimension:     XY
#Bounding box:  xmin: 112.2242 ymin: -44.1284 xmax: 153.8942 ymax: -8.8798
#Geodetic CRS:  GRS 1980(IUGG, 1980)

#read aus: 
#Simple feature collection with 8 features and 1 field
#Geometry type: MULTIPOLYGON
#Dimension:     XY
#Bounding box:  xmin: 112.9211 ymin: -43.63192 xmax: 153.6389 ymax: -9.229614
#Geodetic CRS:  WGS 84
#  adm                       geometry
#1 NSW MULTIPOLYGON (((141.0029 -3...
#2 Vic MULTIPOLYGON (((140.9659 -3...
#3 QLD MULTIPOLYGON (((153.5522 -2...
#4  SA MULTIPOLYGON (((129.0013 -2...
#5  WA MULTIPOLYGON (((129.0012 -1...
##6 Tas MULTIPOLYGON (((148.3096 -4...
#7  NT MULTIPOLYGON (((137.9962 -1...
#8 ACT MULTIPOLYGON (((149.1253 -3...



##Mapping polygons

```{r}
tm_shape(shelf) +
  tm_polygons()
```
#Remember we can make a thematic map by layering it up just as we do for plots in ggplot2. Here we have indicated the shape of our map (shelf) and we have added a command bbox = sdat to expand the extent of the map so it depicts all of our copepod data points. We then add the shape of Australia (aus) on top of the shelf, and finally our copepod data (sdat) in the form of points using tm_dots().

```{r}
tm_shape(shelf, bbox = sdat) +
  tm_polygons() +
  tm_shape(aus) +
  tm_polygons() +
  tm_shape(sdat) +
  tm_dots()
```

#exploring T_map
#quick way to style tmap you could use tmap_style("beaver"). This is a similar function to ggplot


```{r}
vignette('tmap-getstarted')
```





```{r}
library(tidyr)
shelf <- mutate(shelf,)
```


#Playing with tmap

```{r}
library(terra)
```
# raster data use these packages: terra, raster, and stars
```{r}
rsst <- rast('data-for-course/spatial-data/MeanAVHRRSST/MeanAVHRRSST.grd')
```

NOw plot the raster
```{r}
plot(rsst)
```
# notethe colour scale isnt that approirate for temp. Green where temp is hot and red where temp is cold. would rather much do one with breks in the colors, for color blind friendly'

```{r}
tm_shape(rsst) +
  tm_raster(palette = "-RdBu", title = "SST")
```

#note the - infront of palette name, which makes the re warmer temperature by revefsing the colour palette.

#Layering raters and points inTmap

#overlayed our copopod sampling points on the map of temperature
```{r}
tm_shape(rsst) +
  tm_raster(palette = "-RdBu", title = "SST") +
  tm_shape(sdat) +
  tm_dots(col = "richness_raw", 
          palette = "Greys",
          title = "Speices #") +
  tm_compass()
```


















